---
title: "Final Report"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(table1)
```

From rubric: The report is clearly written and sufficiently introduces the dataset of choice, the question of interest, the approach taken, and the results.

# Background

Climate change is very prevalent today and affects every inhabited region in the world. Evidence of climate change is shown by rising sea levels, increased extreme weather, and warmer surface and ocean temperatures. Regarding global surface temperatures, NOAA’s 2021 Annual Climate Report states that they increase annually and are projected to continue increasing. Due to the uneven naturs of warming, average temperatures will reflect the average global temperature, not the temperature in a specific geographic location. We wnat to be able to visualize and communicate data that we have gathered to various stakeholders. 

Our final project aims to increase the ease of investigators to access data and statistical methods. We have created an app that allows people to visualize the change in average temperature historically, as well as the predicted temperature changes for that country out 5 years.

# Data

The data are collected from UC-Berkley’s climate lab, spanning 1750-2013. It is publicly available on Kaggle under "Climate Change: Earth Surface Temperature Data". Specifically, we are utilizing the Global Land Temperatures By City (GlobalLandTemperaturesByCity.csv) file. The variables of interest are Average Temperature, Latitude, Longitude, City, Country, and Date. In the 1980's, Average Temperature started being measured by a digital thermometer which may make recent measurements more accurate, but reflect a cooling bias relative to older measures.

### Data preview:

```{r, include=FALSE}
library(tidyverse)
setwd("/home/gtiernon/biostat625/project")
env_data <- read.csv("GlobalLandTemperaturesByCity.csv") %>% 
  select(-X)
```

```{r}
head(env_data)
dim(env_data)
```

DOES A TABLE 1 MAKE SENSE HERE?? I think we could take some randomly sampled countries?

```{r}
smallNumObs = env_data[which(env_data$Country %in% sample(unique(env_data$Country), 10, replace = FALSE)),]
```



# Research Aim

Our goal is to communicate statistical analysis of global surface temperatures to policy makers or curious citizens via a user-friendly Shiny application. 

# Analysis


# Results

Link to shiny app with maybe a description of trends we observed. 


